---
title: "Bayesian Analysis Task III"
author: "Leonel Fernando Nabaza Ruibal"
date: "`r format(Sys.Date(), '%d/%m/%Y')`"
output:
  pdf_document:
    toc: true
    number_sections: true
geometry: margin=2.5cm
fontsize: 12pt
lang: eng
header-includes:
  - \\usepackage[english]{babel}
  - \\usepackage{amsmath,amssymb,amsthm}
  - \\usepackage{booktabs}
  - \\usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  fig.align = 'center'
)
set.seed(1234)
```

# Wording and data

We want to study whether the final grades in the *Bayesian Analysis* course are the same for boys and girls.

The observed data are:

```{r data}
boys  <- c(9.6, 7.0, 5.0, 8.0, 8.4, 6.4)
girls <- c(6.1, 9.1, 8.8, 5.7, 8.9, 6.1, 6.5)

df <- data.frame(
  grade = c(boys, girls),
  sex   = c(rep(1, length(boys)), rep(0, length(girls))),
  group = c(rep("boys", length(boys)), rep("girls", length(girls)))
)

knitr::kable(df, col.names = c("grade", "sex", "group"), booktabs = TRUE)
```

The statement proposes the model
\[
y_i \mid \beta_0, \beta_1, \sigma \sim \mathcal N(\beta_0 + \beta_1\, \text{sex}_i, \sigma),
\]
where \(\text{sex}_i = 1\) for boys and \(\text{sex}_i = 0\) for girls.

# Initial exploration

```{r initial-summary}
summary_by_group <- aggregate(grade ~ group, data = df, function(x) c(n = length(x), mean = mean(x), sd = sd(x)))
summary_by_group
```

```{r data-boxplot}
library(ggplot2)

ggplot(df, aes(x = group, y = grade)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.1, height = 0) +
  labs(title = "Grades by group", x = "Group", y = "Grade")
```

At first glance, the sample means are very similar, but the dispersion does not appear to be exactly the same in both groups. Moreover, grades are bounded between 0 and 10, so a pure normal model does not strictly respect the support of the variable.

# a) Full Bayesian model

## Model proposed in the statement

The linear model in the statement can be written as
\[
y_i \mid \beta_0, \beta_1, \sigma \overset{ind}{\sim} \mathcal N(\beta_0 + \beta_1 x_i, \sigma), \qquad x_i \in \{0,1\}.
\]

Here:

- \(\beta_0\) is the mean grade for girls.
- \(\beta_0 + \beta_1\) is the mean grade for boys.
- \(\beta_1\) is the difference in means, boys minus girls.
- \(\sigma\) is the common standard deviation for both groups.

A reasonable choice of weakly informative priors is
\[
\beta_0 \sim \mathcal N(7, 2^2), \qquad
\beta_1 \sim \mathcal N(0, 2^2), \qquad
\sigma \sim \text{Half-Normal}(0, 2).
\]

### Justification of the priors

- Grades are on a 0--10 scale, so centering \(\beta_0\) around 7 is plausible and still weakly informative.
- For \(\beta_1\), centering at 0 expresses that *a priori* we do not favor either group having better grades.
- The half-normal prior for \(\sigma\) guarantees positivity and avoids excessively heavy tails in such a small sample.

## Prior plots

```{r prior-plots-statement-model}
par(mfrow = c(1, 3))
curve(dnorm(x, mean = 7, sd = 2), from = -1, to = 12,
      main = expression(paste("Prior for ", beta[0])), xlab = expression(beta[0]), ylab = "density")
curve(dnorm(x, mean = 0, sd = 2), from = -6, to = 6,
      main = expression(paste("Prior for ", beta[1])), xlab = expression(beta[1]), ylab = "density")
xs <- seq(0, 6, length.out = 500)
plot(xs, 2*dnorm(xs, mean = 0, sd = 2), type = "l",
     main = expression(paste("Prior for ", sigma)), xlab = expression(sigma), ylab = "density")
par(mfrow = c(1, 1))
```

# b) Discussion of the model and alternative proposal

The model in the statement is useful as a starting point, but it has two important limitations:

1. **Common variance**. It imposes the same standard deviation for boys and girls. With so few observations, this restriction may be too strong.
2. **Unbounded support**. The normal distribution allows values below 0 and above 10, which does not perfectly match the nature of grades.

## Proposed alternative model

I propose using a model with **different means and different standard deviations** for each group:
\[
y_{g,j} \mid \mu_g, \sigma_g \sim \mathcal N(\mu_g, \sigma_g),
\qquad
y_{b,k} \mid \mu_b, \sigma_b \sim \mathcal N(\mu_b, \sigma_b).
\]

With weakly informative priors:
\[
\mu_g \sim \mathcal N(7, 2^2), \qquad
\mu_b \sim \mathcal N(7, 2^2), \qquad
\sigma_g, \sigma_b \sim \text{Half-Normal}(0, 2).
\]

This model remains simple, but it is more flexible than the original one. If we wanted to be even more realistic with the support \([0,10]\), we could use a truncated normal model or a beta model after rescaling the grades to \((0,1)\). However, for this sample size, the normal model with separate variances is a very reasonable and easy option to implement in Stan.

# Chosen model for Stan

## Stan code

Below we use the model with separate means and separate standard deviations. In addition, in `generated quantities` we request:

- the difference in means \(\delta = \mu_b - \mu_g\),
- a posterior predictive draw for a new girl, `y_new_girl`.

```{r write-stan-file, eval=TRUE}
stan_code <- '
data {
  int<lower=1> N_boys;
  int<lower=1> N_girls;
  vector[N_boys] y_boys;
  vector[N_girls] y_girls;
}
parameters {
  real mu_boys;
  real mu_girls;
  real<lower=0> sigma_boys;
  real<lower=0> sigma_girls;
}
model {
  // priors
  mu_boys ~ normal(7, 2);
  mu_girls ~ normal(7, 2);
  sigma_boys ~ normal(0, 2);
  sigma_girls ~ normal(0, 2);

  // likelihood
  y_boys ~ normal(mu_boys, sigma_boys);
  y_girls ~ normal(mu_girls, sigma_girls);
}
generated quantities {
  real delta;
  real y_new_girl;
  delta = mu_boys - mu_girls;
  y_new_girl = normal_rng(mu_girls, sigma_girls);
}
'
writeLines(stan_code, "bayesian_grades_model.stan")
cat(stan_code)
```

## Fitting with `rstan`

```{r stan-fit, eval=TRUE}
library(rstan)
library(bayesplot)
library(posterior)

rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

stan_data <- list(
  N_boys  = length(boys),
  N_girls = length(girls),
  y_boys  = boys,
  y_girls = girls
)

fit <- stan(
  file = "bayesian_grades_model.stan",
  data = stan_data,
  chains = 4,
  iter = 4000,
  warmup = 2000,
  seed = 1234,
  refresh = 0
)

print(fit, pars = c("mu_boys", "mu_girls", "sigma_boys", "sigma_girls", "delta", "y_new_girl"),
      probs = c(0.025, 0.5, 0.975))
```

> **Note**: the previous chunk is set to `eval=FALSE` so that the document can be opened even if `rstan` is not installed. If you already have it configured, you can change it to `eval=TRUE`.

# c) 95% credible interval for a new female student

We seek the 95% credible interval for the grade of a **new girl** who did not take the exam. Under the chosen model, this corresponds to the posterior predictive distribution of `y_new_girl`.

## Approximate analytical summary

As a quick approximation, using only the girls' data:

```{r girls-stats}
n_g <- length(girls)
mean_g <- mean(girls)
sd_g <- sd(girls)

c(n_girls = n_g, mean_girls = mean_g, sd_girls = sd_g)
```

Under the normal model with the classical noninformative prior, the posterior predictive distribution for a new female observation is a Student \(t\):
\[
y_{\text{new},g} \mid y \sim t_{n_g-1}\left(\bar y_g,\; s_g\sqrt{1+\frac{1}{n_g}}\right).
\]

```{r analytical-interval}
scale_pred <- sd_g * sqrt(1 + 1 / n_g)
ci_analytical <- mean_g + qt(c(0.025, 0.975), df = n_g - 1) * scale_pred
ci_analytical
```

This interval is a useful reference, but the main answer to the exercise should be based on the Bayesian model fitted in Stan.

## Interval from Stan

Once the model has been fitted, the interval is obtained directly from the posterior quantiles of `y_new_girl`:

```{r posterior-interval-stan, eval=FALSE}
post <- rstan::extract(fit)

ci_stan <- quantile(post$y_new_girl, probs = c(0.025, 0.5, 0.975))
ci_stan
```

And we can visualize the posterior predictive distribution:

```{r posterior-predictive-plot, eval=FALSE}
hist(post$y_new_girl, breaks = 40,
     main = "Posterior predictive distribution for a new girl",
     xlab = "y_new_girl")
abline(v = ci_stan[c(1, 3)], lty = 2)
```

# Final interpretation

## On the equality between boys and girls

With these data, the observed difference between means is small. In the Bayesian model fitted with Stan, the key parameter is
\[
\delta = \mu_b - \mu_g.
\]

To assess whether grades are "equal", it is useful to examine:

- whether 0 lies inside the 95% credible interval for \(\delta\),
- how much posterior mass lies above or below 0,
- and whether a small difference would have practical relevance.

Useful code to summarize this is:

```{r delta-summary, eval=FALSE}
ci_delta <- quantile(post$delta, probs = c(0.025, 0.5, 0.975))
prob_boys_higher <- mean(post$delta > 0)

ci_delta
prob_boys_higher
```

If the interval for \(\delta\) contains 0 comfortably, the evidence in favor of differences between sexes will be weak.

## Compact answer for submission

1. The model in the statement is
   \[
   y_i \sim \mathcal N(\beta_0 + \beta_1 x_i, \sigma),
   \]
   with weakly informative priors such as
   \(\beta_0 \sim \mathcal N(7,2^2)\),
   \(\beta_1 \sim \mathcal N(0,2^2)\),
   \(\sigma \sim \text{Half-Normal}(0,2)\).

2. That model is reasonable, but it can be improved by allowing different variances in each group. Therefore, we propose
   \[
   y_{b,k} \sim \mathcal N(\mu_b, \sigma_b), \qquad
   y_{g,j} \sim \mathcal N(\mu_g, \sigma_g),
   \]
   with weakly informative priors on \(\mu_b, \mu_g, \sigma_b, \sigma_g\).

3. Under this model, the 95% credible interval for the grade of a new female student is obtained from the posterior predictive variable `y_new_girl` generated in Stan:
   \[
   \bigl[q_{0.025}(y_{\text{new},g}),\; q_{0.975}(y_{\text{new},g})\bigr].
   \]
   As a quick analytical approximation using the girls' data only, the interval is:

```{r print-final-interval}
knitr::kable(
  data.frame(
    method = "Analytical t approximation",
    lower = round(ci_analytical[1], 3),
    upper = round(ci_analytical[2], 3)
  ),
  booktabs = TRUE,
  col.names = c("Method", "Lower bound", "Upper bound")
)
```

# Compilation instructions

## Option 1: compile the PDF from RStudio

- Open `bayesian_grades_stan.Rmd`.
- Install packages if needed:

```{r install-note, eval=FALSE}
install.packages(c("rmarkdown", "knitr", "rstan", "bayesplot", "posterior"))
```

- If you do not have LaTeX installed:

```{r tinytex-note, eval=FALSE}
install.packages("tinytex")
tinytex::install_tinytex()
```

- Click **Knit to PDF**.

## Option 2: from the R console

```{r render-note, eval=FALSE}
rmarkdown::render("bayesian_grades_stan.Rmd", output_format = "pdf_document")
```
